Missing The Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-12-02 Manuscript.Rmd

- asterisks denote a weak segway between sentences

@@

Iterative imputation has become the de facto standard to accommodate for missing data. The aim is usually to draw valid inferences, i.e. to get unbiased, confidence-valid estimates that incorporate the effects of the missingness. Such estimates are obtained with iterative imputation by separating the missing data problem from the scientific problem. The missing values are imputed (i.e., filled in) using some sort of algorithm. And subsequently, the scientific model of interest is performed on the completed data. To obtain valid scientific estimates, both the missing data problem and the scientific problem should be considered appropriately. The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.

All inferences rely on convergence of the algorithm, **yet** determining whether an algorithm has converged is not trivial.* There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms. A widely accepted practice is visual inspection of the algorithm, **but** identifying convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. Therefore, a quantitative diagnostic method to identify convergence would be preferred.

It is challenging to arrive upon a single point at which convergence has been reached.* Since the aim is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged. Because of this property, it may be more desirable to focus on *non*-convergence. Fortunately, there are non-convergence identifiers for other iterative algorithms, **although** the validity of these identifiers has not been systematically evaluated on imputation algorithms.

In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms. We evaluate whether these methods are able to cover the extent of the non-convergence, **but** we also investigate the relation between non-convergence and the validity of the inferences. We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a case study.

@@

<!-- The aim of the simulation study is to determine the impact of non-convergence on the validity of statistical inferences, and to assess whether non-convergence may be detected using several diagnostic methods. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions (nsim = 1000). -->

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios that are defined by the data generating mechanism. 
<!-- : three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??]. -->

@@

## Methods

@@

As recommended by @buur18, our performance measures are bias, confidence interval width, and coverage rate of the estimands ($\S$ 2.5.2)[or just of the regression estimate?? otherwise add the ciw and cov of r2 too!!].

@@

Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
<!-- The next four plots show the autocorrelations in each parameter plotted against the number of iterations. -->

@@

Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
<!-- The next four plots show the $\widehat{R}$ in the different parameters. -->

@@

<!-- Update with different CI computation, see <https://stefvanbuuren.name/fimd/sec-evaluation.html#sec:quantifyingbias> -->

@@

<!-- Magnitude of the bias in $r^2$ perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for $p=0.95$, MNAR, but stable from it=6. -->

@@

<!-- # Check if heuristic for AC works -->

<!-- ```{r} -->

<!-- load("Data/diagnostics.Rdata") -->

<!-- load("Data/raw_outcomes.Rdata") -->

<!-- sim_nr <- 3 -->

<!-- outcomes[[sim_nr]] %>% filter(mech == "MCAR", p == 0.95) %>% ggplot(aes(x=it, y = est)) + geom_line() -->

<!-- convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.qhat)) + geom_line() -->

<!-- convergence_diagnostics %>% filter(sim == sim_nr, mech == "MCAR", p ==0.95) %>% ggplot(aes(x=it, y = ac.max.lambda)) + geom_line() -->

<!-- ``` -->

>>>>

- asterisks denote a weak 'segway' between sentences

@@

Iterative imputation has become the de facto standard to accommodate for missing data. The aim is usually to draw valid inferences, i.e. to get unbiased, confidence-valid estimates that incorporate the effects of the missingness. Such estimates are obtained with iterative imputation by separating the missing data problem from the scientific problem. The missing values are imputed (i.e., filled in) using some sort of algorithm. And subsequently, the scientific model of interest is performed on the completed data. To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered. The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.

All inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial. There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms. A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. Therefore, a quantitative diagnostic method to assess convergence would be preferred.*

It is challenging to arrive upon a single point at which convergence has been reached.* Since the aim is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged. Because of this property, it may be more desirable to focus on *non*-convergence. Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.

In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms. We evaluate whether these methods are able to cover the extent of the non-convergence, and we also investigate the relation between non-convergence and the validity of the inferences. We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a case study.

@@

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence, we terminate the algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios that are defined by the data generating mechanism.

@@

## Diagnostic methods

@@

As recommended by @buur18, our performance measures are bias, confidence interval width, and coverage rate of the estimands ($\S$ 2.5.2).
<!-- [or just of the regression estimate?? otherwise add the ciw and cov of r2 too!!]. -->

@@

Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).

@@

Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).

@@

@@

@@
